---
title: "PA1_template.Rmd"
author: "Bill Lancaster"
date: "May 28, 2016"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

## Fitbit Activity Monitoring Data

#-----------------------------------------------------------------------------------

#### Loading and preprocessing the data
```{r Loading and preprocessing the data}
setwd("~/Coursera/Course 5 Rep/Project_1") #Setting the directory to correct location on computer. 

download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile = "activity.zip", mode="wb")
unzip("activity.zip")
df<-read.table("activity.csv", sep=",", na.strings = "NA", header=T)
```
Data is downloaded into data frame called df. 
For reproducibility purposes the source zip file and the csv file are preserved in the folder. 

#-----------------------------------------------------------------------------------

#### What is mean total number of steps taken per day?
```{r What is mean total number of steps taken per day}
totalsteps <-aggregate(steps ~ date, df, sum) #calulates the total number of steps taken per day
hist(totalsteps$steps, main = paste("Histogram of Steps Taken per Day"))
summary(totalsteps$steps) #show mean and median of steps taken per day. 
```
Note the histogram above and summary statistics contains missing data.

#-----------------------------------------------------------------------------------

#### What is the average daily activity pattern?
```{r What is the average daily activity pattern}
timeinterval<-aggregate(steps ~ interval, df, mean)
plot(timeinterval, type="l",main = paste("Avg of Steps Taken per 5-Minute Interval"), xlab="5 Minute Intervals")
abline(h=mean(timeinterval$steps), col="blue", lty = 3)
timeinterval[which(timeinterval$steps == max(timeinterval$steps, na.rm = TRUE)), ]
```
Above is the 5-minute interval, on average across all the days in the dataset

#-----------------------------------------------------------------------------------

#### Imputing missing values, and see if histogram is different?
```{r Imputing missing values}
totalNA <-sum(is.na(df$steps)) 
paste("The total number of NA records in this dataset is", totalNA)

idf<-df #create a data frame to have imputed values

##the strategy for imputing missing data is to use the mean for the 5-minute interval
idf$steps[which(is.na(idf))] <-timeinterval$steps #imputes the mean of the 5-minute interval for NA

## test to compare before and after data
head(df)
head(idf)
tail(df)
tail(idf)


itotalsteps <-aggregate(steps ~ date, idf, sum) #calulates the total number of steps taken per day
hist(itotalsteps$steps, main = paste("Histogram of Steps Taken per Day"))
summary(itotalsteps$steps) #show mean and median of steps taken per day. 
```
There is no difference with the histogram between the data with the NA's and the data with the imputed interval. 

#-----------------------------------------------------------------------------------

####Are there differences in activity patterns between weekdays and weekends?
```{r Difference between weekdays and weekends}

library(chron)
library(lattice)


idfw<-idf
idfw$date <-as.Date(idfw$date) #Convert date from a factor to date format
idfw$dayofwk<-weekdays(idfw$date)
idfw$weekend<-is.weekend(idfw$date) #Creates binary result for wkday vs. wkend


idfwf<-subset(idfw, idfw$weekend=="FALSE") ##this is the weeday data
idwftimeinterval<-aggregate(steps ~ interval, idfwf, mean)
idwftimeinterval$weekday<-"weekday"
idfwt<-subset(idfw, idfw$weekend=="TRUE") ## this is the weekend data
idwttimeinterval<-aggregate(steps ~ interval, idfwt, mean)
idwttimeinterval$weekday<-"weekend"

idwtticomb <-rbind(idwftimeinterval,idwttimeinterval)

xyplot(steps ~ interval | weekday, data = idwtticomb,  main='Fitbit Weekday vs. Weekend', type = "l", xlab = "5 Second Intervals", 
    ylab = "Avg Number of steps", index.cond=list(c(2,1)),layout = c(1, 2))

```









